---
title: "TCGA analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,include=FALSE}
library(TCGAbiolinks) # Download data (version: 2.34.0)
library(dplyr) # Processing data (version: 1.1.4)
library(ggplot2) # Representation of the data (version: 3.5.1)
library(gridExtra) # Representation of the data (version: 2.3)
library(ggplotify) # Representation of the data and save the plot (version: )
library(plotly) # Representation of the data (version: 0.1.2)
library(survival) # Representation of survival (version: 3.8-3)
library(survminer) # Representation of survival (version: 0.5.0)
```

## Information of the project

```{r, echo=FALSE}
# We obtain the information about the data that we are going to use
getProjectSummary("TCGA-LUAD")
```

```{r, include=FALSE}
#We obtain the clinical data of lung cancer
query <- GDCquery(
    project = "TCGA-LUAD", 
    data.category = "Clinical",
    data.type = "Clinical Supplement", 
    data.format = "BCR Biotab"
)
# clinical <- getResults(query)
GDCdownload(query)
clinical <- GDCprepare(query)
# Remove the first two rows with information about the column
clinical_data<-data.frame(clinical$clinical_patient_luad[c(-1,-2),])
```

```{r, include=FALSE}
# Preprocess data
## Are there duplicated patients? 
table(duplicated(clinical_data$bcr_patient_barcode)) # No

## Replace unwanted values with NA across all columns. [Discrepancy] is not removed because it is a relevant diagnosis
unwanted_values <- c('[Not Available]', '[Not Evaluated]', '[Unknown]', '[Not Applicable]')
clinical_data <- clinical_data %>%
  mutate(across(everything(), ~ifelse(. %in% unwanted_values, NA, .)))


## Filter the columns that are completely NA
clinical_data_filter_col <- clinical_data[, colSums(!is.na(clinical_data) & clinical_data != "") > 0]
clinical_data_filter_col$age_at_initial_pathologic_diagnosis <- as.numeric(clinical_data_filter_col$age_at_initial_pathologic_diagnosis)
```

## Show all the columns that has not relevant information to analyze (empty)

```{r,echo=FALSE}
names(clinical_data[!names(clinical_data) %in% names(clinical_data_filter_col)])
```
```{r,warning=FALSE,message=FALSE}
## Create the individual plots of age, Vital status and gender
plot_age <- ggplot(clinical_data_filter_col, aes(x = age_at_initial_pathologic_diagnosis)) +
  geom_histogram(binwidth = 10, color = "black", fill = "#7AC5CD", alpha = 0.7) +
  labs(title = "Age Distribution of LUAD Patients", x = "Age", y = "Abs. Frequency") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    axis.text = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5)
  )

plot_vital_status <- ggplot(clinical_data_filter_col, aes(x = vital_status)) +
  geom_bar(fill = c("#66CDAA", "#CD1076"), color = "black") +
  labs(title = "Vital Status of LUAD Patients", x = "Vital Status", y = "Total") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    axis.text = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5)
  )

plot_gender <- ggplot(clinical_data_filter_col, aes(x = gender)) +
  geom_bar(fill = c("#9A32CD", "#1874CD"), color = "black") +
  labs(title = "Gender Distribution of LUAD Patients", x = "Gender", y = "Total") +
  theme_minimal() + 
  theme(
    axis.title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    title = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5),
    axis.text = element_text(size = 12, face = "bold", angle = 0, hjust = 0.5)
  )

# Arrange the plots in a grid
grid_plot <- as.ggplot(grid.arrange(plot_age, plot_vital_status, plot_gender, ncol = 2))
ggsave('./Grid_plots.png',plot = grid_plot,width = 2200,height = 1600,units = 'px')
```

```{r,warning=FALSE,message=FALSE}
ggplot(clinical_data_filter_col, aes(x = age_at_initial_pathologic_diagnosis, fill = vital_status)) +
  geom_histogram(binwidth = 10, color='black') +
  labs(title = "Age Distribution Stratified by Gender and Vital Status", x = "Age", y = "Total") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 14, face = "bold", angle = 0, hjust = 0.5),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12)
  ) +
  facet_wrap(~ gender) +  # Facet by gender
  scale_fill_manual(values = c("#66CDAA", "#CD1076"))
```

```{r,warning=FALSE,message=FALSE}
# First step, filter data with NA in the column ajcc_pathologic_tumor_stage
clinical_data_violin <- clinical_data_filter_col %>% 
  filter(!is.na(ajcc_pathologic_tumor_stage))
# Violin plot to represent the age of the patients along AJCC tumor stage
ggplot(clinical_data_violin, aes(x = ajcc_pathologic_tumor_stage, y = age_at_initial_pathologic_diagnosis)) +
  geom_violin(fill = "lightblue", color = "black", alpha = 0.7) +
  geom_boxplot(width = 0.2, color = "black", alpha = 0.5, show.legend = FALSE) + 
  geom_jitter(aes(color = ajcc_pathologic_tumor_stage), width = 0.2, alpha = 0.6, size = 2) +
  labs(title = "Age Distribution by AJCC Pathologic Tumor Stage", x = "AJCC Pathologic Tumor Stage", y = "Age") +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 12)
  )
```
```{r,warning=FALSE,message=FALSE}
data <- table(clinical_data_filter_col$anatomic_organ_subdivision)

# Convert the regions and the absolute frequency into a dataframe
df <- data.frame(
  Region = names(data),
  Count = as.numeric(data)
)

# Pie chart of the distribution of Lung Regions with plotly (dinamic representation)
plot_ly(df, 
                 labels = ~Region,
                 values = ~Count,
                 type = 'pie',
                 textinfo = 'label+percent',
                 hoverinfo = 'label+percent', 
                 marker = list(colors = c('#66CDAA', '#CD1076', '#1E90FF', '#FFD700', '#8A2BE2'))) %>%
  layout(
    title = "Distribution of Lung Regions", 
    showlegend = TRUE 
  )
```

# Survival analysis

```{r, echo=FALSE,warning=FALSE}
# Ensure columns are numeric
clinical_data_filter_col$death_days_to <- as.numeric(clinical_data_filter_col$death_days_to)
clinical_data_filter_col$last_contact_days_to <- as.numeric(clinical_data_filter_col$last_contact_days_to)

# clinical_data_filter_col vital_status, convert to numeric (e.g., 1 for "Dead", 0 for "Alive")
clinical_data_filter_col$vital_status <- ifelse(clinical_data_filter_col$vital_status == "Dead", 1, 0)

# Create survival_time column
clinical_data_filter_col$survival_time <- ifelse(clinical_data_filter_col$vital_status == 1, 
                                                 clinical_data_filter_col$death_days_to, 
                                                 clinical_data_filter_col$last_contact_days_to)
clinical_data_survival <- clinical_data_filter_col %>%
  filter(!is.na(survival_time) & !is.na(ajcc_pathologic_tumor_stage))
  
```

## Kaplan-Meier survival curves

### Tumor stage (survival distributions between the groups are significantly different)

```{r}
fit <- survfit(Surv(survival_time, vital_status) ~ ajcc_pathologic_tumor_stage, data = clinical_data_survival)
ggsurvplot(fit, data = clinical_data_survival, pval = TRUE, risk.table = FALSE,
           legend = c(0.93, 0.6),
           legend.labs = c("Discrepancy","Stage I","Stage IA","Stage IB","Stage II","Stage IIA","Stage IIB", "Stage IIIA", "Stage IIIB", "Stage IV"),
           legend.title = element_blank()
) 
```

```{r}
ggsurvplot(fit, data = clinical_data_survival, risk.table = TRUE,
           legend.labs = c("Discrepancy","Stage I","Stage IA","Stage IB","Stage II","Stage IIA","Stage IIB", "Stage IIIA", "Stage IIIB", "Stage IV"),
           legend.title = element_blank(),
           risk.table.height = 1
) 
```


### *EGFR* mutation (survival distributions are the same)

```{r}
fit_egfr <- survfit(Surv(survival_time, vital_status) ~ egfr_mutation_status, data = clinical_data_survival)
ggsurvplot(fit_egfr, data = clinical_data_survival, pval = TRUE)
```

### *KRAS* mutation (survival distributions are the same)

```{r}
fit_kras <- survfit(Surv(survival_time, vital_status) ~ kras_mutation_found, data = clinical_data_survival)
ggsurvplot(fit_kras, data = clinical_data_survival, pval = TRUE)
```

## Cox Proportional Hazards Model

```{r,warning=FALSE}
# Adjust the model
cox <- coxph(Surv(survival_time, vital_status) ~ age_at_initial_pathologic_diagnosis + ajcc_pathologic_tumor_stage + tobacco_smoking_pack_years_smoked, data = clinical_data_survival)
# summary(cox)
```

```{r}
# Obtain the results, create a dataframe for the graphic and filter
summary_cox <- summary(cox)
hr_data <- data.frame(
  Variable = rownames(summary_cox$coefficients), 
  HR = exp(summary_cox$coefficients[, "coef"]),  
  lower_ci = exp(summary_cox$conf.int[, "lower .95"]), 
  upper_ci = exp(summary_cox$conf.int[, "upper .95"]),  
  p_value = summary_cox$coefficients[, "Pr(>|z|)"] 
)
hr_data_filtered <- hr_data %>% filter(p_value < 0.05)
```


```{r}
# Forest plot (Only with relevant variables with significative effect on Death)
ggplot(hr_data_filtered, aes(x = HR, y = Variable)) +
  geom_point(aes(color = p_value < 0.05), size = 4) + 
  geom_errorbarh(aes(xmin = lower_ci, xmax = upper_ci), height = 0.2) +
  scale_x_log10() + 
  labs(
    title = "Hazard Ratios (HR) de Supervivencia con Regresión de Cox",
    x = "Hazard Ratio (HR)",
    y = "Variables"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 12),
    title = element_text(size = 16, face = "bold"),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1)
  ) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") 
```


